\documentclass[11pt]{article}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{url}

\lstdefinelanguage{pbcc}{
morekeywords={transform, if, else, for, from, to, using, case, do, where,
              return, or, in_training, secondary, primary, priority, generator,
              template, assert, while, config, MatrixRegion1D, MatrixRegion2D,
              MatrixRegion3D, cell, region, row, col},
sensitive=true,
morecomment=[l]{\#},
morecomment=[l]{//},
morestring=[b]',
morestring=[b]",
}
\lstset{language=pbcc}
\lstset{numbers=left}
\lstset{tabsize=2}
\lstset{commentstyle=\textit}
\lstset{literate=
{REFTransform}{{\ref{sec:transform}}}{3}
{REFRule}     {{\ref{sec:rule}}}{3}
{REFBody}     {{\ref{sec:body}}}{3}
}

\begin{document}


\title{PetaBricks Language Specification}
\author{Jason Ansel}
\maketitle
\tableofcontents

\newpage
\section{Syntax}

\subsection{Minimal PetaBricks Program}

Figure~\ref{ex1} is an example of a (near) minimal PetaBricks program. It
copies the matrix A to the matrix B.

\begin{figure}[h]
\begin{lstlisting}
transform Copy2D                                                 REFTransform
to B[w,h]                                                        REFTransform
from A[w,h]                                                      REFTransform
{
  to(B.cell(x,y) b)                                              REFRule
  from(A.cell(x,y) a)                                            REFRule  
  {                                          
    b=a;                                                         REFBody
  }                                         
}
\end{lstlisting}
This defines the transform Copy2D, which can be invoked as follows:

\begin{lstlisting}
MatrixRegion2D a,b;
...
Copy2D(a,b);
\end{lstlisting}
  \caption{
  A minimal PetaBricks program.  This invocation will result in the rule,
  defined on lines~5-9, being invoked on every pair of elements in $a$
  and $b$ in an undefined order.
  \label{ex1}
  }
\end{figure}


\subsection{Transform Structure}

The high level structure of a transform is as follows.

\begin{lstlisting}
Transform Header  [REFTransform]
{
  Rule Header     [REFRule] 
  {                       
    Rule Body     [REFBody]
  }                      
  
  Rule Header     [REFRule] 
  {                       
    Rule Body     [REFBody]
  }                                         

  ...
}
\end{lstlisting}

\subsection{Transform Header}
\label{sec:transform}

The header of a transform consists of an unordered, whitespace separated
list of clauses.  The minimal set of clauses are TRANSFORM, FROM, and TO.
All of the allowed clauses are defined as follows:

\begin{figure}[h]
\begin{lstlisting}
transform Sum1Da
to Total
from A[n]
{
  //Compute the total sequentially
  to(Total t) from(A a) {
    t=0;
    for(int i=0; i<n; ++i)
      t+=a.cell(i); 
  }
  
  //Divided and conquer
  to(Total t) from(A.region(0,n/2) left, A.region(n/2,n) right) {
    t=Sum1Da(left)+Sum1Da(right);
  }
}
\end{lstlisting}
  \caption{
  Sum all of the elements in an array.
  \label{ex2}
  }
\end{figure}

\begin{figure}[h]
\begin{lstlisting}
transform Sum1Db
to Total
using Partial[n/2]
from A[n]
{
  //sum two neighboring elements
  to(Partial.cell(i) p) from(A.cell(2*i) a1, A.cell(2*i+1) a2){
    p=a1+a2;
  }

  //lower priority rule to handle odd sized arrays
  secondary to(Partial.cell(i) p) from(A.cell(2*i) a1){
    p=a1;
  }

  //recursively call self to finish summing
  to(Total t) from(Partial p){
    Sum1Db(t,p);
  }

  //optionally skip Partial and just compute the output directly
  //this serves as the base case for the recursive calls
  to(Total t) from(A a){
    Sum1Da(t,a);
  }
}
\end{lstlisting}
  \caption{
  Sum all of the elements in an array, by making an intermediate array of size n/2 or by calling Sum1Da.
  \label{ex3}
  }
\end{figure}

\subsubsection{TRANSFORM Clause}

\begin{tabular}{| l | l |}
\hline
\bf Required & Yes                    \\
\bf Available In Version & 1.0        \\
\bf Syntax & \tt transform IDENTIFIER[\ref{IDENTIFIER}] \\
\bf Example & \tt transform Foo       \\
\bf Used in & \tt Figure~\ref{ex1}      \\
\hline
\end{tabular}

~

\noindent
Specifies the externally visible name of the transform.

\subsubsection{FROM Clause}

\begin{tabular}{| l | l |}
\hline
\bf Required & Yes                    \\
\bf Available In Version & 1.0        \\
\bf Syntax & \tt from MATRIXDEFLIST[\ref{MATRIXDEFLIST}] \\
\bf Example & \tt from A[w,h], B[w]     \\
\bf Used in & \tt Figure~\ref{ex2}      \\
\hline
\end{tabular}

~

\noindent
\noindent Defines the inputs of the transform.  The sizes of the outputs
are given as expressions containing unbound variables.  The compiler and
runtime must find a valid assignment for these unbound variables to match
the sizes of the inputs given.  The values assigned to these variables are
visible both in rule headers and in rule bodies.


\subsubsection{TO Clause}

\begin{tabular}{| l | l |}
\hline
\bf Required & Yes                    \\
\bf Available In Version & 1.0        \\
\bf Syntax & \tt to MATRIXDEFLIST[\ref{MATRIXDEFLIST}]     \\
\bf Example & \tt to C[w,h], D[w]     \\
\bf Used in & \tt Figure~\ref{ex2}      \\
\hline
\end{tabular}

~

\noindent Defines the outputs of the transform.  The sizes of the outputs
are given as expressions containing unbound variables that are set based on the size
of the inputs.

\subsubsection{USING Clause}

\begin{tabular}{| l | l |}
\hline
\bf Required & No                          \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt using MATRIXDEFLIST[\ref{MATRIXDEFLIST}]     \\
\bf Example & \tt using E[w,h], F[w]     \\
\bf Used in & \tt Figure~\ref{ex3}      \\
\hline
\end{tabular}

~

\noindent Defines intermediate data used during the transform execution. This
data is temporary and neither an input nor an output.  The sizes are given
as expressions containing unbound variables that are set based on the size
of the inputs.

\subsubsection{GENERATOR Clause}


\begin{tabular}{| l | l |}
\hline
\bf Required & No \\
\bf Available In Version & 2.0            \\
\bf Syntax & \tt generator IDENTIFIERLIST[\ref{IDENTIFIERLIST}]\\
\bf Example & \tt generator Bar\\
\bf Used in & \tt Figure~\ref{ex4}      \\
\hline
\end{tabular}

~

\noindent Specifies the name of another transform (or list of transforms)
that should be used to generate inputs to this transform during training.
The outputs of the other transforms are paired left to right with the inputs
of this transform.  The other transforms inputs are seeded with random data
of the desired input size.

If this clause is not used, transforms will be given random data during
training.

\begin{figure}[h]
\begin{lstlisting}
transform Permute
to B[n]
from A[n], Indices[n]
generator PermuteGenerator
{
  to(B.cell(i) b) from(A a, Indices.cell(i) idx){
    assert idx < n;
    b=a.cell(idx);
  }
}

//generate random test data for Permute
transform PermuteGenerator 
to A[n], Indices[n]
from A[n]
{
  to(Indices.cell(i) idx) {
    idx=randint(0, n-1);
  }
}
\end{lstlisting}
  \caption{
  Example usage for generator keyword.  Computes $B[i]=A[Indicies[i]]$.
  \label{ex4}
  }
\end{figure}

\subsubsection{TEMPLATE Clause}


\begin{tabular}{| l | l |}
\hline
\bf Required & No \\
\bf Available In Version & 2.0            \\
\bf Syntax & \tt template < IDENTIFIER(INT, INT), ... > \\
\bf Example & \tt template < precision(1,5) > \\
\bf Used in & \tt Figure~\ref{ex5}      \\
\hline
\end{tabular}

~

\noindent Defines a parameter to the transform that requires independent
training for each value.  This causes N transforms to be generated, one for
each possible value.  The first INT specifies the minimum value while the
second specifies the maximum value.  The new transforms can be called with
the syntax: Name$<$ParamValue$>$.

\begin{figure}[h]
\begin{lstlisting}
template < reps(1,5) >
transform Foo
to B[n]
from A[n]
{
  to(B b) from(A a) {
    Bar(b, a);  
    for(int i=1; i<reps; ++i)
      Bar(b, b);
  }
}

MatrixRegion1D a,b;
...
Foo<3>(b, a);
\end{lstlisting}
\caption{
Example usage for template keyword.
\label{ex5}
}
\end{figure}

\subsubsection{Default Values}

\begin{tabular}{| l | l |}
\hline
\bf Required & No \\
\bf Available In Version & 2.0            \\
\bf Example & \tt from IN[h,w]=OUT[h,w], iterations=5 \\
\hline
\end{tabular}

In the {\tt FROM} clause, the programmer may specify default values for transform inputs.
These default values work the same way as default values in {\tt C++}.  This provides an easy way for defining in place transforms.

\subsection{Rule Header}
\label{sec:rule}

The header for a rule consists of an unordered, whitespace-separated list
of clauses.  The meanings of these clauses are defined below.  The only
required clause is TO.

\subsubsection{TO Clause}
\label{TO}
\begin{tabular}{| l | l |}
\hline
\bf Required & Yes                          \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt to( MATRIXARGLIST[\ref{MATRIXARGLIST}] ) \\
\bf Example & \tt to( A.cell(x,y) a, B.region(0, 0, x, y) b)\\
\hline
\end{tabular}

~

\noindent Specifies the outputs for the rule.

The coordinates in the TO and FROM clause must correspond to an affine
transformation of each other.

\subsubsection{FROM Clause}
\label{FROM}
\begin{tabular}{| l | l |}
\hline
\bf Required & No                          \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt from( MATRIXARGLIST[\ref{MATRIXARGLIST}] ) \\
\bf Example & \tt from( A.cell(x,y) a, B.region(0, 0, x, y) b)\\
\hline
\end{tabular}

~

\noindent Specifies the inputs for the rule.

The coordinates in the TO and FROM clause must correspond to an affine
transformation of each other.

\subsubsection{USING Clause}
\label{USING}
\begin{tabular}{| l | l |}
\hline
\bf Required & No \\
\bf Available In Version & 2.0             \\
\bf Syntax & \tt using( MATRIXDEFLIST[\ref{MATRIXDEFLIST}] ) \\
\bf Example & \tt using( T1[10], T2[w+h, w+h] ) \\
\hline
\end{tabular}

~

\noindent Defines temporary buffers required by the body of the rule.
PetaBricks will automatically allocate and reuse these temporary buffers.

\subsubsection{HINT Clause}
\begin{tabular}{| l | l |}
\hline
\bf Required & No \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt hint( EXPRESSION[\ref{EXPRESSION}] )   \\
\bf Example & \tt hint( n )           \\
\hline
\end{tabular}

~

\noindent Provide a hint to to autotuner that the decision to use this
rule should be based on comparing the given EXPRESSION to some constant.
Note that as of version 2.0 this language feature should no longer be needed
except in special cases.


\subsubsection{WHERE Clause}
\begin{tabular}{| l | l |}
\hline
\bf Required & No                          \\
\bf Available In Version & 2.0             \\
\bf Syntax & \tt where EXPRESSION[\ref{EXPRESSION}] \\
\bf Example & \tt where x<10               \\
\hline
\end{tabular}

~

\noindent Limits the range where this rule may be applied.

\subsubsection{PRIORITY Clause}
\begin{tabular}{| l | l |}
\hline
\bf Required & No                          \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt priority(int)             \\
\bf Example & \tt priority(1)              \\
\hline
\end{tabular}

~

\noindent Specify the priority for the given rule.  This is a clean way for
specifying handling of corner cases.  For a given cell, only rule with a
minimum priority may be applied.  The default priority is $1$.

\subsubsection{PRIMARY Keyword}
\begin{tabular}{| l | l |}
\hline
\bf Required & No                          \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt primary             \\
\hline
\end{tabular}

~

\noindent Equivalent to {\tt priority(0)}

\subsubsection{SECONDARY Keyword}
\begin{tabular}{| l | l |}
\hline
\bf Required & No                          \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt secondary \\
\hline
\end{tabular}

~

\noindent Equivalent to {\tt priority(2)}

\subsubsection{ROTATABLE Keyword}
\begin{tabular}{| l | l |}
\hline
\bf Required & No                          \\
\bf Available In Version & 1.0             \\
\bf Syntax & \tt rotatable \\
\hline
\end{tabular}

~

TODO: Define this more precisely.

\noindent Syntactic sugar for defining four rules with the coordinate 
system rotated by $\frac \pi 2$ in each consecutive one.

\subsubsection{Return-style}

\begin{tabular}{| l | l |}
\hline
\bf Required & No \\
\bf Available In Version & 1.0        \\
\hline
\end{tabular}

~

PetaBricks allows a shorter version of rule headers that eliminates
the need for a TO clause.  It is similar to function style return syntax.
 
\begin{minipage}{\linewidth}
\begin{lstlisting}
B.cell(x,y) from(A.cell(x,y) a) {
  return a;
}
\end{lstlisting}
\end{minipage}
Is desugared to:

\begin{minipage}{\linewidth}
\begin{lstlisting}
to(B.cell(x,y) _rv) from(A.cell(x,y) a) {
  _rv=a;
  return;
}
\end{lstlisting}
\end{minipage}

\subsubsection{OR Keyword}

\begin{tabular}{| l | l |}
\hline
\bf Required & No \\
\bf Available In Version & 2.0        \\
\hline
\end{tabular}

~

PetaBricks provides an ``OR'' keyword to eliminate the need for redundant 
rule headers.  ``OR'' simply means repeat the last rule header. For example:
 
\begin{minipage}{\linewidth}
\begin{lstlisting}
to(B b) from(A a) {
  Foo(b,a);
} or {
  Bar(b,a);
}
\end{lstlisting}
\end{minipage}
Is desugared to:

\begin{minipage}{\linewidth}
\begin{lstlisting}
to(B b) from(A a) {
  Foo(b,a);
}
to(B b) from(A a) {
  Bar(b,a);
}
\end{lstlisting}
\end{minipage}

\subsection{Rule Body}
\label{sec:body}

Rule bodies are standard {\tt C++} code with some preprocessing performed.
All variables defined in the corresponding rule header are visible from the 
rule body.

\subsubsection{MatrixRegion Class}
Inputs an output matrix regions are of the template type {\tt
MatrixRegion<N,T>}, where $N$ is the number of dimensions and $T$ is the
element type.  The typedefs: {\tt MatrixRegion0D}, {\tt MatrixRegion1D},
{\tt MatrixRegion2D}, ... are provided to refer to output matrix regions of each
dimensionality.  Additionally the typedefs {\tt ConstMatrixRegion0D}, {\tt
ConstMatrixRegion1D}, {\tt ConstMatrixRegion2D}, ... refer to matrix
regions with read only data and are used for inputs to a rule.

This class is by reference, so using it never copies data. To make a copy,
one must explicitly call {\tt MatrixRegion::allocate} and then copy.
MatrixRegion supports numerous operations not discussed here.  The API
documentation for this class can be found here:

\noindent
\url{http://kleptocracy.csail.mit.edu/pbapi/d9/db0/classpetabricks_1_1MatrixRegion.html}

\subsubsection{Configuration/Tunable Parameters}

The user may specify configuration parameters with:
\begin{verbatim}
config(name, initial, min, max);
\end{verbatim}
in the file header.  Both min and max are optional and default to 0 and
MAX\_INT.  These parameters are stored in the program configion file, but are
{\bf not} autotuned.  The value may be changed by hand, with {\tt configtool}
(included in the scripts folder), and at runtime with {\tt name.setValue(int)}.
The current value can be accessed with either {\tt name} or {\tt name.value()}.

The user may specify tunable parameters with:
\begin{verbatim}
tunable(name, initial, min, max);
\end{verbatim}
in the file header.  This is the same as a {\tt config} except that autotuner 
will attempt to determine an optimal value automatically.

\subsubsection{Training Runs}


Some transforms require a training phase to determine configuration parameters
such as the number of iterations.  The autotuner supports this through the
API function {\tt PetabricksRuntime::isTrainingRun()}, or the conditional macro
{\tt in\_training}.  On the first execution with a given configuration,
calling this method will return {\tt true} and as a side effect the timing
results will be discarded and the test re-run.  On subsequent executions
this method will return {\tt false} and have no side effects.

\begin{figure}[h]
\begin{lstlisting}
transform Foo
to B[n]
from A[n]
{
  to(B b) from(A a) {
    config(reps);
    in_training {
      // experimentally determine how many reps are needed
      // for convergence in the current configuration
      reps = 0;
      while(! isConverged(b,a) ){
        reps++;
        Bar(b, b, a);
      }
    }else{
      for(int i=1; i<reps; ++i){
        Bar(b, b, a);
      }
    }
  }
}
\end{lstlisting}
\caption{
Example usage for training runs
\label{ex6}
}
\end{figure}

\subsubsection{Clean Abort}

Transforms can abort execution (in the event of an invalid configuration or
input) using the API function {\tt PetabricksRuntime::abort()}.  During training
this will cleanly end the current trail and remove the source configuration
from consideration.

\subsection{Definitions}
\subsubsection{MATRIXDEFLIST}
\label{MATRIXDEFLIST}
 
A comma separated list of {\tt MATRIXDEF[\ref{MATRIXDEF}]}.

\subsubsection{MATRIXDEF}
\label{MATRIXDEF}

Declares a new matrix of a given (symbolic) size.

Of the format: {\tt NAME} [ {\tt DIMENSIONS} ] 

Where {\tt NAME} is an identifier and {\tt DIMENSIONS} is a comma separated
list of list of symbolic expressions.

\subsubsection{MATRIXARGLIST}
\label{MATRIXARGLIST}

A comma separated list of {\tt MATRIXARG[\ref{MATRIXARG}]}.

\subsubsection{MATRIXARG}
\label{MATRIXARG}

Declares an argument to a rule that references a matrix defined by a MATRIXDEF.

Of the format: {\tt MATRIX.REGIONSPEC } {\tt NAME}

Where MATRIX is name of the matrix the data is from.  REGIONSPEC is the part
of the matrix to get and NAME is the identifier defined by this statement.

\subsubsection{REGIONSPEC}

One of:

\begin{itemize}
\item \tt cell(EXPRESSION...)
\item \tt region(EXPRESSION...)
\item \tt row(EXPRESSION)
\item \tt col(EXPRESSION)
\item \tt all()

\end{itemize}

\subsubsection{IDENTIFIERLIST}
\label{IDENTIFIERLIST}

A comma separated list of {\tt IDENTIFIER[\ref{IDENTIFIER}]}.

\subsubsection{IDENTIFIER}
\label{IDENTIFIER}

A name matching the regular expression: {\tt [a-zA-Z\_][a-zA-Z0-9\_]*}

\subsubsection{EXPRESSION}
\label{EXPRESSION}

An algebraic expression consisting of identifiers, integers, floats, and the operators: [*/()+-]

Unless stated otherwise, expressions must correspond to affine transformation
of the output coordinates.

\section{Semantics}

\subsection{Semantics for Rule Execution}

The header (defined Sec.~\ref{sec:rule}) of each rule contains a set of
free variables used to reference an explicit list input and output regions.
A rule may be {\em instantiated} by binding these free variables to values
determined by the compiler and runtime.  The compiler may bind these free
variables to any value such that:
\begin{itemize}
\item All values are non-negative integers.
\item All regions referenced exist (i.e. not out of bounds access).
\item All input regions referenced have already been computed.
\item Any WHERE clauses evaluate to true.
\end{itemize}

Once the inputs and outputs are bound to an explicit set of cells, rule
execution follows a semantics and syntax similar to that of {\tt C++} with
the additional features defined in Section~\ref{sec:body}.


\subsection{Semantics for Transform Execution}

The semantics for the execution of a transform revolve around executing
rules in parallel.  This section defines a internal state of an executing
transform and an operational semantics over this internal state for how a
running transform may progress from an initial state to a completion state.

The following definitions are required:
\begin{itemize}
\item
Let $r[\alpha]$ represent rule $r$ instantiated with free variables
bound to the set of definitions $\alpha$.
\item
Let $\mbox{depends}(r[\alpha])$ be the set of cells $r[\alpha]$
depends on (all the regions specified in the FROM[\ref{FROM}] clause when
free variables are bound using the mapping $\alpha$).
\item
Let $\mbox{provides}(r[\alpha])$ be the set of cells $r[\alpha]$
outputs on (all the regions specified in the TO[\ref{TO}] clause when free
variables are bound using the mapping $\alpha_j$).
\item Let $T$ be a static reference of a transform, consisting of a set of rules.
\item Let $R$ be a set of rule instances $r_0[\alpha_0] ... r_n[\alpha_n]$
\item Let $D$ be a set of cells.  Each element in this contains both the $x,y$ coordinate and the matrix in which the cell belongs.
\item Let $\langle T, R, D_0, D_1, D_2 \rangle$ represent the internal state for a running transform where:
  \begin{itemize}
  \item $T$ is the transform being executed.
  \item $R$ are the rules that may currently executing in parallel.  (Initially $\epsilon$)
  \item $D_0$ are cells already computed.  (Initially those defined by the FROM clause.)
  \item $D_1$ are cells that can be computed, but need not be. (Initially those defined by the USING clause.) 
  \item $D_2$ are cells that must be computed.  (Initially those defined by the TO clause.)
  \end{itemize}
\end{itemize}

The following transition rules govern the progression of the internal state
of a transform during execution.  This semantics intentionally gives a lot of
freedom to the PetaBricks runtime.  Specifically,  the order of computation
is defined only as producers-before-consumers, data may be computed more
than once, and unfinished computations may be arbitrarily discarded.

~

\noindent
{\it
Transform initialization:}

~

\begin{tabular}{c}
Transform $T$ is called with inputs $D_0$ and outputs $D_2$.
\\
$D_1$ is the intermediate (USING) data required by $T$.
\\\hline
$\rightarrow \langle T, \epsilon, D_0, D_1, D_2 \rangle$ 
\end{tabular}

~

\noindent
{\it
Rule execution begins:}

~

\begin{tabular}{c}
$r \in T$ \\
$\alpha$ is a legal assignment for free variables in $r$ \\
depends$(r[\alpha]) \subset D_0$  
\\\hline
$
\langle T, R, D_0, D_1, D_2 \rangle
\rightarrow
\langle T, R \cup \{r[\alpha]\}, D_0, D_1, D_2 \rangle$ 
\end{tabular}

~

\noindent
{\it
Rule execution ends:}

~

\begin{tabular}{c}
$r[\alpha]$ has finished executing
\\
$D_0' = D_0 \cup {\mbox{provides}(r[\alpha])}$
\\
$D_1 = D_1' \cup {\mbox{provides}(r[\alpha])}$
\\
$D_2 = D_2' \cup {\mbox{provides}(r[\alpha])}$
\\\hline
$
\langle T, R \cup \{r[\alpha]\}, D_0, D_1, D_2 \rangle
\rightarrow
\langle T, R, D_0', D_1', D_2' \rangle$
\end{tabular}

~

\noindent
{\it
Rule execution aborted:}

~

\begin{tabular}{c}
$
\langle T, R \cup \{r[\alpha]\}, D_0, D_1, D_2 \rangle
\rightarrow
\langle T, R, D_0, D_1, D_2 \rangle$
\end{tabular}


~

\noindent
{\it
Transform completion:}

~

\begin{tabular}{c}
$R=\epsilon$
\\
$D_2=\epsilon$
\\\hline
$
\langle T, R, D_0, D_1, D_2 \rangle
\rightarrow$
\end{tabular}

\end{document}


